---
title: "hdmax2 application for multivariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_multivariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction 

The `hdmax2` package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary or categorial as well as multivariate exposure.

In this vignette, we treat example with a two variables exposure.

To install the latest version of `hdmax2`, use the github repository 

```{r package installation}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated data for 100 samples and 500 potential mediators.

For this example we use a two variables exposure (continuous and binary), a continuous outcome and two adjustment factors (age and gender). The matrix of potential mediators is an extract of real methylation data (TCGA PDAC samples).


We define the $K$ number of estimated latent factors by performing a PCA on potential mediators matrix. The scree plot criterion looks for the “elbow” in the curve and selects all components just before the line flattens out, $K$ is chosen with this procedure.


```{r variable definition}
simu_data = hdmax2::simu_data

## Outcomes
Y = simu_data$Y_continuous

## High dimension mediators
M = simu_data$M

## Covariates
age = simu_data$age
gender = simu_data$gender

```

```{r pca for latent factor}
## Number of Latent factor estimation
pc <- prcomp(M)
plot((pc$sdev^2/sum(pc$sdev^2))[1:10],
     type = "b",
     xlab = 'Principal Component',
     ylab = "Explained variance")

K=5 #pca conclusion : it is better to select too many factors that too few
```


## Use case  *multivariate exposure* and *continuous outcome*




### STEP 1: Run association studies

The `run_AS` function is applied to identify significant effects of exposure $X$ on potential mediators in $M$ matrix, and significant effect of potential mediators on outcome $Y$. 
The `run_AS` function takes as inputs: $X$ for exposure, $Y$ for outcome, $M$ for the potential mediators matrix, and $K$ for the number of latent factors defined earlier.  Adjustment factors (covariates) can be included. 

- $X$ must be a data frame with  $n$ rows 2 colums (for each exposure variable).
- $Y$ can be a vector or a matrix with $n$ rows and 1 column.
- $M$ must be a matrix with $n$ rows and $p$ columns (potential mediators).
- $K$ must be integer.
- *covar* must be numeric (data frame or matrix) with $n$ rows.


```{r step 1, message=FALSE}
## multivariate Exposures (continuous + binary)
X = data.frame(X1 = simu_data$X_continuous, X2 = simu_data$X_binary)
covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X = X ,
                              Y = Y,
                              M = M,
                              K = K,
                              covar = covar)


```


### Selection of a subset of mediators

Step 1 provides an object containing:

- Results from the two association studies ($P$-values, fscores, zscores),

- Latent factors estimation matrix $U$ ($p$ rows and $K$ columns) from first regression, 

- `run_AS` function's inputs, 

- and *max-squared* test $P$-values results. 


These $P$-values are used in the selection of mediators for the user's chosen method. 

Numerous selection methods are available, such as FDR control. Also in the context of methylation data, it's feasible to aggregate methylation probes. Refer to the *helper_functions* vignette for assistance FDR control and AMR researching procedure.

In this instance, we opted for the top ten most significant $P$-values from the *max-squared* test as an example.


```{r mediators selection}
## Selecting top 10 mediators
mediators_top10 = M[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function `estimate_effect` estimate the individual indirect effect of mediators, but also overall effects of selected mediators.
The function `estimate_effect` takes as inputs, step 1 object and selected mediators matrix $M^s$  from chosen selection method apply on *max-squared* test $P$-values.


```{r  step2, message=FALSE}

hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = mediators_top10)
```



This step use `mediation::mediate` function to obtain several effects estimation:

- ACME Average Causal Mediation Effect

- PM Proportion Mediate

- TE total effect

- ADE Average Direct Effect

This step also compute Overall effects : 

- OIE (Indirect effect)

- ODE (Direct Effect)

- OTE (Total Effect)

And regression effects size.

### Vizualisation of results

We propose a set of plots which including:

- Mediators ACME Forest plot

- Mediators PM Forest plot

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r plot results, fig.height=10, fig.width=10}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2, plot_type= "all_plot")
```



   - **A** Estimates of indirect effect (ACME) and **B** proportions of mediated effect (PM) for the top 10 mediators. The effect estimate is represented by a dot and its 95\% CI by the bar. Symbols correspond to the significance cut off of 5\%  (square for p-value $\geq 0.05$, circle p-value $< 0.05$). Colors correspond to the sign of the effect (green for estimated effect $\leq 0$ , red for estimated effect $> 0$). 
   
   - **C** Effect sizes of Overall Direct Effect (ODE), Overall Indirect Effect (OIE) and Overall Total Effect (OTE). Error bars correspond to standard deviation (ODE and OTE) or confidence interval (OIE). 
    
   - **D** Indirect effect sizes for the selected mediators. Black corresponds to the ACME, violet to the effect of exposure $X$ on mediator $M$, and blue corresponds to the effect of mediator $M$ on outcome $Y$.

In the `plot_hdmax2` function it is possible to produce the 4-plots set or each individual plot with *plot_type* argument.

For each exposure variable a set of plot is performed.


```{r, eval = FALSE}
sessionInfo()
```

