---
title: "hdmax2 application for univariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_univariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction 

The {\tt hdmax2} package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, as well as multivariate exposomes (which includes categorical variables). 

In this vignette, we treat the univariate example.

To install the latest version of **hdmax2**, use the github repository 

```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated 100 samples and 500 potential mediators, with various type of exposures (continuous or binary) and outcomes(continuous or binary).


```{r}

simu_data = hdmax2::simu_data

## Exposures
X_matrix = as.matrix(simu_data$X_binary)

## Outcomes
Y_matrix = as.matrix(simu_data$Y_continuous)

## High dimension mediators
M_matrix = as.matrix(simu_data$M)

## Covariates
age = as.matrix(simu_data$age)
gender = as.matrix(simu_data$gender)

## Number of Latent factore estimation
pc <- prcomp(M_matrix)
plot(pc$sdev[1:10],
     xlab = 'Principal Component',
     ylab = "Explained variance")


K=8 #pca conclusion : it is better to select too many factors that too few
```


## First use case  *binary exposure* and *continuous outcome*

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposuire $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r}

 X_matrix = as.matrix(as.numeric(X_matrix))
 K = K
 X_type= "univariate"
 Y_type = "continuous"
 covar = cbind(age, gender)

hdmax2_step1 = run_AS(X_matrix = X_matrix ,
                Y_matrix = Y_matrix,
                 M_matrix = M_matrix, 
                 K = K,
                 X_type = X_type, # indicates the type of Exposure
                 Y_type = Y_type,  # indicates the type of Outcome
                 covar = covar)

```

### Selection of a subset of mediators

In this example, we will consider the 10 first selected mediators.


```{r}
## Selecting top 10 mediators
## top 10 selecting mediators
mediators_top10 = M_matrix[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2
The function **estimate_effect** estimate the individual indirect effect of mediators.

```{r}
object = hdmax2_step1
m = mediators_top10

hdmax2_step2 = estimate_effect(object = object,
                                    m = m)

```


### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r}
library(ggplot2)

plot_univariate(hdmax2_step2, N_med = 10)

```



```{r}
sessionInfo()
```