---
title: "hdmax2 application for multivariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_multivariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction 

The {\tt hdmax2} package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, or categorical, as well as multivariate exposomes. Binary variables are converted to 0s and 1s and treated as univariate variables. 

In this vignette, we treat the univariate example.

To install the latest version of **hdmax2**, use the github repository 
```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated 100 samples and 500 potential mediators, with various type of exposures (continuous or binary) and outcomes(continuous or binary).


```{r}

simu_data = hdmax2::simu_data

## Exposures
X_matrix = simu_data$X_categorial
X_matrix = droplevels(X_matrix)
X_matrix = mltools::one_hot(data.table::as.data.table(X_matrix))
X_matrix = as.matrix(X_matrix)

## Outcomes
Y_matrix = as.matrix(simu_data$Y_continuous)

## High dimension mediators
M = as.matrix(simu_data$M)

## Covariates
age = as.matrix(simu_data$age)
gender = as.matrix(simu_data$gender)

## Number of Latent factore estimation

res_pca = FactoMineR::PCA(M, graph = F)
factoextra::fviz_screeplot(res_pca, ncp=10)

K=5 #pca conclusion : it is better to select too many factors that too few
```


## First use case  *binary exposure* and *continuous outcome*

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposuire $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r}




  M_matrix = M

  X_type= "multivariate"
  detailed = TRUE
  Y_type = "continuous"
  covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X_matrix = X_matrix ,
                 Y_matrix = Y_matrix,
                 M_matrix = M,
                 K = K,
                 X_type = X_type, # indicates the type of Exposure
                 Y_type = Y_type,  # indicates the type of Outcome
                 detailed = detailed,  # indicates the exposure is multivariate
                 covar = covar)

# 
# # max2 test results
 head(hdmax2_step1$max2_pvalues)

```

### Selection of a subset of mediators

In this example, we will consider the 10 first selected mediators.


```{r}
##Selecting mediators
## top 10 selecting mediators
mediators_top10 = M[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2
The function **estimate_effect** estimate the individual indirect effect of mediators.

```{r}
object = hdmax2_step1
 m = mediators_top10
  boots = 100
  sims = 100
## selected mediators effects estimation



 hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                     m = mediators_top10,
                                      boots = 100,
                                sims = 100,
                                is.categorial = TRUE)


##The parameter sims is the number of Monte Carlo draws for nonparametric bootstrap or quasi-Bayesian approximation. 10000 is recommended, but we used 100 for the example.
```

Estimation of **average causal mediation effect (ACME)** is the indirect effect. 

Other effect estimated include :

   - Average direct effect (ADE) 

   - Proportion mediated (PM) 

   - Total effect (TE)

```{r}
# ACME, ADE, PM, TE estimation with CI
#head(hdmax2_step2$ACME)
#head(hdmax2_step2$ADE)
#head(hdmax2_step2$PM)
#head(hdmax2_step2$TE)
```


### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r}
library(ggplot2)
## mediators ACME plots 
object = hdmax2_step2
hdmax2::plot_hdmax2(hdmax2_step2, N_med = 10)

```



```{r}
sessionInfo()
```