---
title: "hdmax2 application for multivariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_multivariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```



## Introduction 

The `hdmax2` package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, as well as multivariate exposomes (which includes categorical variables). 

In this vignette, we treat the multivariate example, with a categorial explanatory variables

To install the latest version of **hdmax2**, use the github repository 
```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated 100 samples and 500 potential mediators, with various type of exposures (continuous, binary or categorial) and outcomes(continuous or binary).


```{r}
simu_data = hdmax2::simu_data

## Exposures
X_matrix = simu_data$X_categorial

## Outcomes
Y_matrix = simu_data$Y_continuous

## High dimension mediators
M_matrix = simu_data$M

## Covariates
age = simu_data$age
gender = simu_data$gender

## Number of Latent factor estimation

pc <- prcomp(M_matrix)
plot(pc$sdev^2/sum(pc$sdev)[1:10],
     xlab = 'Principal Component',
     ylab = "Explained variance")

K=5 #pca conclusion : it is better to select too many factors that too few
```


## First use case  *categorial exposure* and *continuous outcome*

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r}


detailed = FALSE
covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X = X_matrix,
                              Y = Y_matrix,
                              M = M_matrix,
                              K = K,
                              detailed = detailed,  # indicates the exposure is multivariate
                              covar = covar)

 
# # max2 test results
head(hdmax2_step1$max2_pvalues)

```

### Selection of a subset of mediators

In this example, we will consider the 10 first selected mediators.


```{r}
## Selecting top 10 mediators
mediators_top10 = M_matrix[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function **estimate_effect** estimate the individual indirect effect of mediators.

```{r}
object = hdmax2_step1
m = mediators_top10
hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = mediators_top10)
```

### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2)

```



## Second use case  *multivariate exposure* and *continuous outcome*


```{r}
simu_data = hdmax2::simu_data
## multivariate Exposures (continuous + binary)
X_matrix = data.frame(X1 = simu_data$X_continuous, X2 = simu_data$X_binary)

## Outcomes
Y_matrix = simu_data$Y_continuous

## High dimension mediators
M_matrix = simu_data$M

## Covariates
age = simu_data$age
gender = simu_data$gender

 K= 5
```

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r}

detailed = FALSE
covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X = X_matrix ,
                              Y = Y_matrix,
                              M = M_matrix,
                              K = K,
                              detailed = detailed,  # indicates the exposure is multivariate
                              covar = covar)

# # max2 test results
head(hdmax2_step1$max2_pvalues)

```

### Selection of a subset of mediators

In this example, we will consider the 10 first selected mediators.


```{r}
## Selecting top 10 mediators
mediators_top10 = M_matrix[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function **estimate_effect** estimate the individual indirect effect of mediators.

```{r}
object = hdmax2_step1
m = mediators_top10
hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = mediators_top10)
```

### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2)

```



```{r}
sessionInfo()
```