---
title: "hdmax2 application for multivariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_multivariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction 

When a statistical association is observed between an external exposure ($X$) and an individual outcome ($Y$), one or more intermediate variables ({\bf M}) (such as gene expression or epigenetic changes) may mediate this effect. Identifying and assessing the mediating role of these variables in the effect of $X$ on $Y$ is crucial for deciphering underlying causal mechanisms in epidemiological and clinical research.

The package  `hdmax2` is dedicated to high-dimensional mediation analyses. For mediation analyses, the necessary data includes one (or more) exposure variable(s), a matrix of potential mediators, and an outcome variable.

The term "high dimensional" refers to the large quantity of potential mediators among which we seek to identify the actual mediators.

The fundamental concept behind HDMAX2 methods is to use a latent factor mixed regression model for estimating unobserved latent factors while conducting high-dimensional association analysis.

The package  `hdmax2` is designed to accept 

- Exposure $X$ consisting of univariate data, which can be continuous, binary or categorial as well as multivariate exposure.

- Outcome $Y$ can be continuous or binary

- Potential mediators $M$ must be continuous.

In this vignette, we treat example with a **two variables** exposure.

To install the latest version of `hdmax2`, use the github repository 

```{r package installation}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated data for 100 samples and 500 potential mediators.

For this example we use a **two variables exposure** (continuous and binary), a **continuous outcome** and **two adjustment factors** (age and gender). The matrix of potential mediators is an extract of real methylation data (TCGA PDAC samples).


We define the $K$ number of estimated latent factors by performing a PCA on potential mediators matrix. The scree plot criterion looks for the “elbow” in the curve and selects all components just before the line flattens out, $K$ is chosen with this procedure.


```{r variable definition}
simu_data = hdmax2::simu_data

## Outcomes
Y = simu_data$Y_continuous

## High dimension mediators
M = simu_data$M2

## Covariates
age = simu_data$age
gender = simu_data$gender


```

```{r pca for latent factor}
## Number of Latent factor estimation
pc <- prcomp(M)
plot((pc$sdev^2/sum(pc$sdev^2))[1:10],
     type = "b",
     xlab = 'Principal Component',
     ylab = "Explained variance")

K=5 #pca conclusion : it is better to select too many factors that too few
```


## Use case  *multivariate exposure* and *continuous outcome*




### STEP 1: Run association studies

The `run_AS` function is applied:

-  First to estimate latent factors with $lfmm$ algorithms.

-  Then to identify significant effects of exposure $X$ on potential mediators in $M$ matrix, and significant effect of potential mediators on outcome $Y$. 

-  And eventually to compute mediation test: *max-squared* test.



The `run_AS` function takes as inputs: 

-  $X$ for exposure: can be a vector, a factor with $n$ (samples) elements or a data frame with  $n$ rows and 1 column.
-  $Y$ for outcome: can be a vector or a matrix with $n$ rows and 1 column.
-  $M$ for the potential mediators: must be a matrix with $n$ rows and $p$ columns (potential mediators).
-  $K$ for the number of latent factors defined earlier: must be integer.
-  Adjustment factors *covar* can be included: must be numeric (data frame or matrix) with $n$ rows.

Here $n$=100 and $p$=500


```{r step 1, message=FALSE}
## multivariate Exposures (continuous + binary)
X = data.frame(X1 = simu_data$X_continuous, X2 = simu_data$X_binary)
covar = cbind(age, gender)
covar = as.data.frame(covar)

hdmax2_step1 = hdmax2::run_AS(X = X ,
                              Y = Y,
                              M = M,
                              K = K,
                              covar = covar)


```




The `run_AS` function provides an object containing:

- Results from the two association studies ($P$-values, fscores, zscores),

- Latent factors estimation matrix $U$ ($p$ rows and $K$ columns) from first regression, 

- `run_AS` function's inputs, 

- And *max-squared* test $P$-values results. 

### Selection of a subset of mediators

*max-squared*  $P$-values are used in the selection of mediators for the user's chosen method. 

Numerous selection methods are available, such as FDR control. Also in the context of methylation data, it's feasible to aggregate methylation probes. Refer to the *helper_functions* vignette for assistance FDR control and AMR researching procedure.

In this instance, we opted for the top ten most significant $P$-values from the *max-squared* test as an example.


```{r mediators selection}
## Selecting top 10 mediators
mediators_top10 = M[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function `estimate_effect` estimate the individual indirect effect of mediators, but also overall effects of selected mediators.

The function `estimate_effect` takes as inputs, step 1 object and selected mediators matrix $M^S$ from chosen selection method apply on *max-squared* test $P$-values.

The function `estimate_effect` use `mediation::mediate` function to obtain several effects estimation with uncertainty:

- ACME Average Causal Mediation Effect: corresponding to the indirect effect

- PM Proportion Mediate: corresponding to the proportion of the total effect that is mediated by the mediator

- TE total effect: which is equal to the sum of direct and indirect effect

- ADE Average Direct Effect: which represents the unmediated effect.

This step also compute Overall effects : 

- OIE (Indirect effect): corresponding to the sum of the indirect effect associated with all mediators.

- ODE (Direct Effect): corresponding to the effect of exposure variables on the outcome variable.

- OTE (Total Effect): corresponding to the effect of exposure variables on the outcome variable when the mediators $M^S$ are included in the model.

In addition, function `estimate_effect` estimates the intermediary effect sizes  $a_j$ and $b_j$ and their standard deviations.


```{r  step2, message=FALSE}

hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = mediators_top10)
```



This step use `mediation::mediate` function to obtain several effects estimation:

- ACME Average Causal Mediation Effect

- PM Proportion Mediate

- TE total effect

- ADE Average Direct Effect

This step also compute Overall effects : 

- OIE (Indirect effect)

- ODE (Direct Effect)

- OTE (Total Effect)

And regression effects size.

### Vizualisation of results

We propose a set of plots which including:

- Mediators ACME Forest plot

- Mediators PM Forest plot

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r plot results, fig.height=10, fig.width=10}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2, plot_type= "all_plot")
```



   - **A** Estimates of indirect effect (ACME) and **B** proportions of mediated effect (PM) for the top 10 mediators. The effect estimate is represented by a dot and its 95\% CI by the bar. Symbols correspond to the significance cut off of 5\%  (square for p-value $\geq 0.05$, circle p-value $< 0.05$). Colors correspond to the sign of the effect (green for estimated effect $\leq 0$ , red for estimated effect $> 0$). 
   
   - **C** Effect sizes of Overall Direct Effect (ODE), Overall Indirect Effect (OIE) and Overall Total Effect (OTE). Error bars correspond to standard deviation (ODE and OTE) or confidence interval (OIE). 
    
   - **D** Indirect effect sizes for the selected mediators. Black corresponds to the ACME, violet to the effect of exposure $X$ on mediator $M$, and blue corresponds to the effect of mediator $M$ on outcome $Y$.

In the `plot_hdmax2` function it is possible to produce the 4-plots set or each individual plot with *plot_type* argument.

For each exposure variable a set of plot is performed.


```{r, eval = FALSE}
sessionInfo()
```

