---
title: "hdmax2 application for univariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_univariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction 

The `hdmax2` package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, as well as multivariate exposomes (which includes categorical variables). 

In this vignette, we treat the univariate example.

To install the latest version of `hdmax2`, use the github repository 

```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated 100 samples and 500 potential mediators, with various type of exposures (continuous or binary) and outcomes (continuous or binary).

For this example we use a binary exposure, a continuous outcome and two covariates (age and gender).

We define the K number of estimated latent factors with a PCA and elbow procedure apply to screeplot.


```{r}

simu_data = hdmax2::simu_data

## High dimension mediators
M = simu_data$M

## Exposures
X  = simu_data$X_binary

## Outcomes
Y  = simu_data$Y_continuous


## Covariates
age = simu_data$age
gender = simu_data$gender
covar = cbind(age, gender)

## Number of Latent factore estimation
pc <- prcomp(M)
plot((pc$sdev^2/sum(pc$sdev^2))[1:10],
     type = "b",
     xlab = 'Principal Component',
     ylab = "Explained variance")

K=5 #pca conclusion : it is better to select too many factors that too few
```


## First use case  *binary exposure* and *continuous outcome*

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r}

hdmax2_step1 = hdmax2::run_AS(X = X,
                              Y = Y,
                              M = M, 
                              K = K,
                              covar = covar)

```

### Selection of a subset of mediators

Step 1 provides an object which contains results from the two regression (pvalues, fscores, zscores), latent factor estimation from first regression, function inputs, and more important max-squared test pvalues results. This last result allow to select mediators which user's choosen method. For this example we selected top ten of more significant pvalues from max-squared test.



```{r}
## Selecting top 10 mediators
## top 10 selecting mediators
mediators_top10 = M[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2
The function **estimate_effect** estimate the individual indirect effect of mediators, but also overall effects of selected mediators.

```{r}
object = hdmax2_step1
m = mediators_top10

hdmax2_step2 = hdmax2::estimate_effect(object = object,
                                       m = m)

```


### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2)

```
We obtained one set of 4 plots:

   - **A** Estimates of indirect effect (ACME) and **B** proportions of mediated effect (PM) for the top 10 mediators. The effect estimate is represented by a dot and its 95\% CI by the bar. Symbols correspond to the significance cut off of 5\%  (square for p-value $\geq 0.05$, circle p-value $< 0.05$). Colors correspond to the sign of the effect (green for estimated effect $\leq 0$ , red for estimated effect $> 0$). 
   
   - **C** Effect sizes of Overall Direct Effect (ODE), Overall Indirect Effect (OIE) and Overall Total Effect (OTE). Error bars correspond to standard deviation (ODE and OTE) or confidence interval (OIE). 
    
   - **D** Indirect effect sizes for the selected mediators. Black corresponds to the ACME, violet to the effect of exposure X on mediator M in the model $X \sim M$, and blue corresponds to the effect of mediator M on outcome Y in the model $Y \sim M + X$.

##TODO comments about plots


```{r}
sessionInfo()
```