---
title: "hdmax2 application for multivariate exposome"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{hdmax2_multivariate.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```



## Introduction 

The `hdmax2` package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, as well as multivariate exposomes (which includes categorical variables). 

In this vignette, we treat two example of multivariate exposomes:

- the first example, with a categorial explanatory variables

- the second example with a two continuous variables exposure

To install the latest version of **hdmax2**, use the github repository 

```{r, package installation}
#devtools::install_github("bcm-uga/hdmax2")
```


## Simulated dataset

We simulated 100 samples and 500 potential mediators, with various type of exposures (continuous, binary or categorial) and outcomes(continuous or binary).

For the two example we will use a continuous outcome and two covariates (age and gender).

We will define the K number of estimeted latent factors with a PCA and elbow procedure apply to screeplot.


```{r, variable definition}
simu_data = hdmax2::simu_data

## Outcomes
Y_matrix = simu_data$Y_continuous

## High dimension mediators
M_matrix = simu_data$M

## Covariates
age = simu_data$age
gender = simu_data$gender

```

```{r, pca for latent factor}
## Number of Latent factor estimation

pc <- prcomp(M_matrix)
plot(pc$sdev^2/sum(pc$sdev)[1:10],
     xlab = 'Principal Component',
     ylab = "Explained variance")

K=5 #pca conclusion : it is better to select too many factors that too few
```


## First use case  *categorial exposure* and *continuous outcome*

### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included or not, for the example we decided to include two covariates defined above. 
Exposure is categorial , for the example we kept vector with character type that will be transformed in factor inside function,  but users can apply transformation to factor if needed (to order categories).

```{r,step 1}
## Exposures is categorial, character vector
X_matrix = simu_data$X_categorial


detailed = FALSE
covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X = X_matrix,
                              Y = Y_matrix,
                              M = M_matrix,
                              K = K,
                              detailed = detailed,  # indicates the exposure is multivariate
                              covar = covar)

 
# # max2 test results
head(hdmax2_step1$max2_pvalues)

```



### Selection of a subset of mediators

Step 1 provides an object which contains results from the two regression (pvalues, fscores, zscores), latent factor estimation from first regression, function inputs, and more important max-squared test pvalues results. This last result allow to select mediators which user's choosen method. For this example we selected top ten of more significant pvalues from max-squared test.



```{r, mediators selection}
## Selecting top 10 mediators
mediators_top10 = M_matrix[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function **estimate_effect** estimate the individual indirect effect of mediators but also overall effects of selected mediators..

```{r, step2}
object = hdmax2_step1
m = mediators_top10
hdmax2_step2 = hdmax2::estimate_effect(object = object,
                                       m = m)

```

### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r, plot results}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2)


```
We obtained three set, for each category of categorial exposure (minus the reference category), of 4 plots:

   - **A** Estimates of indirect effect (ACME) and 
   
   - **B** proportions of mediated effect (PM) for the top 10 mediators. The effect estimate is represented by a dot and its 95\% CI by the bar. Symbols correspond to the significance cut off of 5\%  (square for p-value $\geq 0.05$, circle p-value $< 0.05$). Colors correspond to the sign of the effect (green for estimated effect $\leq 0$ , red for estimated effect $> 0$). 
   
   - **C** Effect sizes of Overall Direct Effect (ODE), Overall Indirect Effect (OIE) and Overall Total Effect (OTE). Error bars correspond to standard deviation (ODE and OTE) or confidence interval (OIE). 
    
   - **D** Indirect effect sizes for the selected mediators. Black corresponds to the ACME, violet to the effect of exposure X on mediator M in the model $X \sim M$, and blue corresponds to the effect of mediator M on outcome Y in the model $Y \sim M + X$.
   
   

## Second use case  *multivariate exposure* and *continuous outcome*




### STEP 1: Run association studies

The **run_AS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of potential mediators, and the effect of each potential mediators on outcome $Y$. Covariables can be included. 

```{r, ex2 step 1 }
## multivariate Exposures (continuous + binary)
X_matrix = data.frame(X1 = simu_data$X_continuous, X2 = simu_data$X_binary)
detailed = FALSE
covar = cbind(age, gender)

hdmax2_step1 = hdmax2::run_AS(X = X_matrix ,
                              Y = Y_matrix,
                              M = M_matrix,
                              K = K,
                              detailed = detailed,  # indicates the exposure is multivariate
                              covar = covar)

# # max2 test results
head(hdmax2_step1$max2_pvalues)

```

### Selection of a subset of mediators

In this example, we will consider the 10 first selected mediators.


```{r, ex2 mediators selection}
## Selecting top 10 mediators
mediators_top10 = M_matrix[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
head(mediators_top10)

```

### STEP 2 : estimation of the indirect effect

The function **estimate_effect** estimate the individual indirect effect of mediators.

```{r, ex2 step2}
object = hdmax2_step1
m = mediators_top10
hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = mediators_top10)
```

### Plot

We propose a set of plots which include:

- Forest plot of mediators ACME

- Forest plot of mediators PM 

- Comparison of ODE, OIE and OTE

- Mediators effect size representation


```{r, ex2 plot results}
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2)

```

We obtained two set, for each variables of multivariate exposure, of 4 plots:

   - **A** Estimates of indirect effect (ACME) and 
   
   - **B** proportions of mediated effect (PM) for the top 10 mediators. The effect estimate is represented by a dot and its 95\% CI by the bar. Symbols correspond to the significance cut off of 5\%  (square for p-value $\geq 0.05$, circle p-value $< 0.05$). Colors correspond to the sign of the effect (green for estimated effect $\leq 0$ , red for estimated effect $> 0$). 
   
   - **C** Effect sizes of Overall Direct Effect (ODE), Overall Indirect Effect (OIE) and Overall Total Effect (OTE). Error bars correspond to standard deviation (ODE and OTE) or confidence interval (OIE). 
    
   - **D** Indirect effect sizes for the selected mediators. Black corresponds to the ACME, violet to the effect of exposure X on mediator M in the model $X \sim M$, and blue corresponds to the effect of mediator M on outcome Y in the model $Y \sim M + X$.
 


```{r}
sessionInfo()
```