---
title: "hdmax2 helper functions"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{helper_functions.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#options("mc.cores" = 1)

```



## Introduction 

The `hdmax2` package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, or categorical, as well as multivariate exposomes. Binary variables are converted to 0s and 1s and treated as univariate variables. 

In this vignette, we provide a series of helper function to process the data.

To install the latest version of `hdmax2`, use the github repository 
```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## How to analyse agregated methylated regions ?

We simulated 500 samples and 1000 potential DNA methylation mediators, with various a binary exposure (smoking status of mothers) and continuous outcomes (birth weight).

### Identifying aggregated mediator regions (AMR)


Identify Aggregated Methylated regions (AMR) with AMR_search function which uses from the P-values from *max-squared* test compute in `run_AS` function and using a adapted comb-p method (comb-p is a tool that manipulates BED files of possibly irregularly spaced P-values and calculates auto-correlation, combines adjacent P-values, performs false discovery adjustment, finds regions of enrichment and assigns significance to those regions). AMR identification could be useful in Epigenomic Wide Association Studies when single CpG mediation is unsuccessful, or could be complementary analysis.


```{r AMR , message=FALSE}
data = hdmax2::helper_ex

#Artificial reduction of dataset size to pass the github action check when building hdmax2 website
data$methylation = data$methylation[ , 800:1000]
data$annotation = data$annotation[800:1000, ]


pc <- prcomp(data$methylation)
plot((pc$sdev^2/sum(pc$sdev^2))[1:15],
     xlab = 'Principal Component',
     ylab = "Explained variance",
     col = c(rep(1, 3), 2, rep(1, 16)))
# chosen number of dimension
K=5

## run hdmax2 step1
hdmax2_step1 = hdmax2::run_AS(X = data$exposure,
                              Y = data$phenotype,
                              M = data$methylation, 
                              K = K)

##Detecting AMR
seed = 0.6 #Careful to change this parameter when working with real data
res.amr_search = hdmax2::AMR_search(chr = data$annotation$chr,
                                    start = data$annotation$start,
                                    end = data$annotation$end,
                                    pval = hdmax2_step1$max2_pvalues,
                                    cpg = data$annotation$cpg,
                                    seed = seed, 
                                    nCores = 2)

res.amr_search$res

res.arm_build = hdmax2::AMR_build(res.amr_search, 
                                  methylation = data$methylation, 
                                  nb_cpg = 2)

#List of DMR selected
head(res.arm_build$res)

##CpG in the DMR
res.arm_build$CpG_for_each_AMR

```

### Quantifying indirect effects 

Like with single mediators analysis *esimate_effect* function could be use to estimate different effects of AMR. Also `plot_hdmax2` can be applied to step2 results.


```{r step2, message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
## run hdmax2 step2
object = hdmax2_step1
mediators_top10 = data$methylation[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
m = as.matrix(res.arm_build$AMR_mean)
boots = 100
## selected mediators effects estimation

hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = m,
                                       boots = 100)
library(ggplot2)
hdmax2::plot_hdmax2(hdmax2_step2, plot_type= "all_plot")
```

## How to select potential mediators using q-values and FDR control ?

Several methods can be use to select potential mediators from step1 results of our method, in our main use cases we simply select top 10 mediators to simplify the narrative. Among available methods to select mediators from mediation test p-values, we can use FDR (False Discovery Rate) control. Q-value is obtained  from *max-squared* test p-values from step 1 with `fdrtools::fdrtools` function then bounded by chosen threshold. 

```{r FDR, message=FALSE}

data_sim = hdmax2::simu_data
## run hdmax2 step1
hdmax2_step1 = hdmax2::run_AS(X = data_sim$X_binary,
                 Y = data_sim$Y_continuous,
                 M = data_sim$M1, 
                 K = K)

## Select candidate mediator  
qval <- fdrtool::fdrtool(hdmax2_step1$max2_pvalues, statistic = "pvalue", plot = F, verbose = F)$qval
candidate_mediator <- qval[qval<= 0.2] # In this example, we will consider FDR levels <20%.
candidate_mediator

```

## How to transform a categorial variable into an ordered factor?

When categorical variables are used as exposure variables, `hdmax2` uses the first category as a reference (intercept) to calculate the effects associated with the variable's other categories. The functions `HDMAX2::run_AS` and `HDMAX2::estimate_effect` will transform the character vector you have used (if any) into a factor, with an arbitrary ordering of the categories.  If order is important to you, here's a simple way to turn your character vector into a factor and order the categories as you wish. You can then use this variable as input to the hdmax2 functions.


```{r factor, message=FALSE}

# inverse alphabetical order for example
X = data_sim$X_categorial
X = factor(X, ordered = TRUE, levels = c("D","C","B","A"))

```


## How to handle adding an additional set of covariates to the second association study?

It is possible to add a second adjustment factors set in association study between potential mediators and outcome if it makes sense from a biological standpoint. Nevertheless, this additional set of adjustment factors for the second association study must include the adjustment factors from the first association study.

```{r second covar set, message=FALSE}

covar = as.data.frame(data$covariables[,1:2])
covar_sup_reg2 = as.data.frame(data$covariables[,3:4])

hdmax2_step1 = hdmax2::run_AS(X = data$exposure,
                              Y = data$phenotype,
                              M = data$methylation,
                              K = K,
                              covar = covar,
                              covar_sup_reg2 = covar_sup_reg2)



```

```{r}
sessionInfo()
```
