---
title: "hdmax2 helper functions"
author: "Florence Pittion, Magali Richard, Olivier Francois"
date: "January, 2024"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{helper_functions.Rmd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#options("mc.cores" = 1)

```



## Introduction 

The {\tt hdmax2} package is designed to accept exposure $X$ consisting of univariate data, which can be continuous, binary, or categorical, as well as multivariate exposomes. Binary variables are converted to 0s and 1s and treated as univariate variables. 

In this vignette, we provide a series of helper function to process the data.

To install the latest version of **hdmax2**, use the github repository 
```{r}
#devtools::install_github("bcm-uga/hdmax2")
```


## How to analyse agregated methylated regions ?

We simulated 500 samples and 1000 potential DNA methylation mediators, with various a binary exposure (smoking status of mothers) and continuous outcomes (birth weight).

### Identifying aggregated mediator regions (AMR)


Identify Aggregated Methylated regions (AMR) with AMR_search function which uses from the P-values from function max2 and using a adapted comb-p method (comb-p is a tool that manipulates BED files of possibly irregularly spaced P-values and calculates auto-correlation, combines adjacent P-values, performs false discovery adjustment, finds regions of enrichment and assigns significance to those regions).


```{r}
data = hdmax2::helper_ex

#Artificial reduction of dataset size to pass the github action check when building hdmax2 website
data$methylation = data$methylation[ , 800:1000]
data$annotation = data$annotation[800:1000, ]


pc <- prcomp(data$methylation)
plot((pc$sdev^2/sum(pc$sdev))[1:15],
     xlab = 'Principal Component',
     ylab = "Explained variance",
     col = c(rep(1, 3), 2, rep(1, 16)))
# choosen number of dimension
K=5

## run hdmax2 step1
hdmax2_step1 = hdmax2::run_AS(X = data$exposure,
                              Y = data$phenotype,
                              M = data$methylation, 
                              K = K)

##Detecting AMR
chr = data$annotation$chr
start = data$annotation$start
end = data$annotation$end
pval = hdmax2_step1$max2_pvalues
cpg = data$annotation$cpg

res.amr_search = hdmax2::AMR_search(chr = data$annotation$chr,
                                    start = data$annotation$start,
                                    end = data$annotation$end,
                                    pval = hdmax2_step1$max2_pvalues,
                                    cpg = data$annotation$cpg,
                                    seed = 0.6, #Careful to change this parameter when working with real data
                                    nCores = 2)

res.amr_search$res

res.arm_build = hdmax2::AMR_build(res.amr_search, 
                                  methylation = data$methylation, 
                                  nb_cpg = 2)

#List of DMR selected
head(res.arm_build$res)

##CpG in the DMR
res.arm_build$CpG_for_each_AMR

```

### Quantifying indirect effects 

Like in single mediators analysis *esimate_effect* function could be use to estimate different effects of AMR.


```{r}
## run hdmax2 step2

object = hdmax2_step1
mediators_top10 = data$methylation[,names(sort(hdmax2_step1$max2_pvalues)[1:10])]
m = as.matrix(res.arm_build$AMR_mean)
boots = 100
sims = 100
## selected mediators effects estimation

hdmax2_step2 = hdmax2::estimate_effect(object = hdmax2_step1,
                                       m = m,
                                       boots = 100,
                                       sims = 100)


#hdmax2::plot_hdmax2(hdmax2_step2)

```

## How to select potential mediators using q=values and FDR control ?

Several methods can be use to select potential mediators from step1 results for our methods, in our main use case we simply select top 10 mediators but we can use multiple test adapted threshold like FDR control (False Discovery Rate)

```{r}

data = hdmax2::simu_data
## run hdmax2 step1
hdmax2_step1 = hdmax2::run_AS(X = data$X_continuous,
                 Y = data$Y_continuous,
                 M = data$M, 
                 K = K)

## Select candidate mediator  
qval <- fdrtool::fdrtool(hdmax2_step1$max2_pvalues, statistic = "pvalue", plot = F, verbose = F)$qval
candidate_mediator <- qval[qval<= 0.8] # In this example, we will consider FDR levels <80%.
candidate_mediator

```

## How to transform a categorial variable using one-hot encoding ?

When categorial variables are used in our models, we must go through design matrices, and for that, one-hot encoding function is needed.

```{r}
simu_data = hdmax2::simu_data

X_matrix = simu_data$X_categorial
X_matrix = as.factor(X_matrix)
X_matrix = droplevels(X_matrix)
X_matrix = mltools::one_hot(data.table::as.data.table(X_matrix))
X_matrix = as.matrix(X_matrix)
```


```{r}
sessionInfo()
```