---
title: "Overview of R Package hdmax2"
author: "Basile Jumentier, Claire-Cécile Barrot, Olivier Francois"
date: "September 8, 2022"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

***

**Summary:**

***

## Introduction 

HDMAX2 is an efficient algorithm for high dimensional mediation analysis using max-squared tests, considering CpGs and aggregated mediator regions.
HDMAX2 is structured in three main steps:
Step 1. Evaluating associations between exposure, mediators and outcome, using a latent factor mixed models (LFMMs) to estimate the effects of exposure $X$ on a matrix $M$ of CpG markers, and the effect of each marker on outcome $Y$.

$$
 M=Xa^T +U_1V_1^T +E_1
$$
$a$ the vector of effect sizes of exposure on DNAm levels,
$U$ the matrix formed of K latent factors estimated simultaneously with a,
$V$ loadings associated with the latent factors,
$E_1$ the matrix of residual errors.

$$
Y=Xc+Mb^T +U_2V_2^T +E_2 
$$
$c$ the direct effect of exposure on outcome,
$b$ the effect sizes of DNAm levels on outcome,
$U_2$ the latent factors from a latent factor regression model,
$V_2$ the corresponding loadings,
$E_2$ the a matrix of errors.

Step 2. Identifying potential CpG mediators and aggregated mediator regions, combining the significance values $P_x$ and $P_y$ to compute a P-value for each CpG.
$$
P = max(P_x, P_y)^2
$$
Quantifying indirect effects with single and multiple mediators.


To install the latest version of **hdmax2**, use the github repository. The library Biostrings from Bioconductor is requierd. 
```{r}
devtools::install_github("bcm-uga/hdmax2")
```

***
## Starting with hdmax2

The hdmax2 package can be loaded as follows. 
```{r}
library(hdmax2)
```

In this tutorial, we present an example of mediation analysis. The examples file include randomized human DNA methylation values, with simulated treatment (exposure), simulated phenotype (outcome), and simulated covariables.  

```{r}
load("example_hdmax2.data")
```
***

## Evaluating associations between exposure, mediators and outcome

In order to estimate how many latent factors can be infered from the methylation matrix, we perform a principal component analysis (PCA) using the **prcomp** function as follows. 

```{r}
pc <- prcomp(methylation)
plot(pc$sdev[1:15]^2,
     xlab = 'Principal Component',
     ylab = "Explained variance",
     col = c(rep(1, 3), 2, rep(1, 16)))
```

The screeplot indicates around 4 main components in the data. We will use $K = 4$ latent factors in subsequent analyses.

The **mEWAS** function is applied to estimate the effects of exposure $X$ on a matrix $M$ of CpG markers, and the effect of each marker on outcome $Y$. It uses the covariables matrix $conf$ and $K$ latent factors.

```{r}
##Epigenome Wide Association Study with both exposure and outcome
res.mEWAS <- mEWAS(X = exposure,
                   Y = phenotype,
                   M = methylation,
                   K = 4,
                   conf = covariables) 
```

The function estimates the effects of $X$ and $Y$ on the matrix $M$. It also estimate the effect of the latent factors.

```{r}
##P-values of X and Y
pvalues <- res.mEWAS$pValue

##Scores
score <- res.mEWAS$score

##scores matrix for the K latent factors
U <- res.mEWAS$U

```

The function **mEWAS** also computes the Genomic Inflation Factor (gif), which expresses the deviation of the distribution of the observed test statistic compared to the distribution of the expected test statistic.

```{r}
##Genomic inflation factor for X and Y
res.mEWAS$gif
```

In case of highly stratified population (high gif), it allows to calibrate P-values without excess false positive rate.

```{r}
##Calibrated P-values of X and Y
pvalues.cal <- res.mEWAS$calibrated.pvalue

##Calibrated scores
score_latent.cal <- res.mEWAS$calibrated.score2
```
***
## CpG mediators

**Identifying potential CpG mediators**

Combining the significance values $pvalues$ with the function **max2**.
This rejects the null-hypothesis that either the effect of exposure on DNAm or the effect of DNAm on outcome is null.

```{r}
cpg_max2 <- max2(pval1 = pvalues[, 1], pval2 = pvalues[, 2], diagnostic.plot = T)
```
The function **max2** computes a P-value and estimate the Q-value for every CpG, and quantify the local false discovery rate (FDR) parameter $eta0$.

```{r}
##local FDR parameter
cpg_max2$eta0
```

In this example, we will consider FDR levels <10%.


```{r}
##Selecting CpG
cpg <- cpg_max2$qval[cpg_max2$qval<= 0.1]
cpg
```

**Quantifying indirect effects of single mediator**

The function **wrap_mediation** estimate the effects of CpG.
```{r}
effect_cpg <- wrap_mediation(qval = cpg_max2$qval,
                            X = exposure,
                            Y = phenotype,
                            M = methylation,
                            U = res.mEWAS$U,
                            sims = 100, FDR = 0.1)

##The parameter sims is the number of Monte Carlo draws for nonparametric bootstrap or quasi-Bayesian approximation. 10000 is recommended, but we used 100 for the example.


```

Estimation of the average causal mediation effect (ACME) is the indirect effect. Other effect estimated include the average direct effect (ADE), the proportion mediated (PM), and the total effect.

```{r}
cpg.acme <- effect_cpg$ACME
cpg.ade <- effect_cpg$ADE
cpg.pm <- effect_cpg$PM
cpg.te <- effect_cpg$TE
#Regression
reg.xm <- effect_cpg$xm
reg.my <- effect_cpg$my
```
***
## Identifying aggregated mediator regions (AMR)

**Identifying AMR**

The function **AMR_search** will identify aggregated methylated regions (AMR) using the max2 P-values. It will compute the P-value and the FDR for each AMR detected.

```{r}
##Detecting AMR
res.amr_search <- AMR_search(chr = annotation$chr,
                   start = annotation$start,
                   end = annotation$end,
                   pval = cpg_max2$pval,
                   cpg = annotation$cpg)
res.amr_search$res
```
The function **AMR_build** will build, for each AMR, a vector with the detail of the AMR, and the first components of each PCA for each DMR. Only AMR with a minimal number of CpG are considered.

```{r}
res.arm_build <- AMR_build(res.amr_search, methylation = methylation, nb_cpg = 2)
#List of DMR selected
res.arm_build$res
```

```{r}
##CpG in the DMR
res.arm_build$CpG_for_each_AMR
```

**Quantifying indirect effects**

The function **wrap_mediation_AMR** is similar to the function **wrap_mediation**.
```{r}
effect_amr <- wrap_mediation_AMR(X = exposure,
                            Y = phenotype,
                            AMR = res.arm_build$AMR_mean,
                            U = res.mEWAS$U,
                            sims = 100)
```

```{r}
##Indirect effect
effect_amr$ACME
```
```{r}
##Direct effect
effect_amr$ADE
```
```{r}
##Peoportion mediated
effect_amr$PM
```
```{r}
##Total effect
effect_amr$TE
```
```{r}
##regressions of X on the DMR
effect_amr$xm
```
```{r}
##regressions of the DMR on Y
effect_amr$my
```




```{r}
#Overall_Indirect_Effect
oie <- est_oie(X = exposure,
               Y = phenotype,
               C = covariables,
               m = cbind(cpg, U),
               boots = 150)

```



